---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Load libraries

```{r}
library(tidyverse)
library(ggplot2)
library(tidyr)
library(dplyr)

```
## Loading and preprocessing the data

Let's read the csv and unpack it from the zip file. Assign it to a df

```{r}
activity_raw_df <-read.csv(unz('activity.zip','activity.csv'),header = T)
head(activity_raw_df)
```

Let's get some stats on the raw data.

```{r}
summary(activity_raw_df)
```
I notice that the date is in character format. I think this would be better in a date format for creating plots, and just for analytic purposes.

```{r}
library(lubridate)

activity_raw_df$date <- ymd(activity_raw_df$date)

str(activity_raw_df)
  
```
Looks like the date field is now in date f <-format. Yeah!

## What is mean total number of steps taken per day?

First, let's do the histogram of steps per day.

```{r}

# first select data and organize it.

raw_data_steps <- activity_raw_df %>% 
      select (date, steps) %>%
      group_by(date) %>% 
      summarise(total_daily_steps = sum(steps))
      
  
    ggplot(data = raw_data_steps,mapping = aes(x=date, y=total_daily_steps, fill=total_daily_steps))+
              geom_bar(stat="identity")+
              scale_fill_continuous(low="light blue", high = "blue")+
              theme(legend.position = "none",                                          axis.title.x=element_blank(),axis.title.y=element_blank())+
              ggtitle("Total Steps by Day")+
              labs(caption = "Raw Data Plot")

```
Now to calculate the mean and median of steps per day. Just using summary on the df.

```{r}

summary(raw_data_steps)

```
# What is the average daily activity pattern?

Create a time series plot of the average daily steps (y axis) of the 5 minute interval (x axis) average all days.

```{r}
# Get the data in the right format for the plot.

raw_data_interval <- activity_raw_df %>% 
      select (interval, date, steps) %>%
      na.omit()  %>% 
      group_by(interval ) %>% 
      summarise(avg_daily_steps = mean(steps))

ggplot(data = raw_data_interval,mapping = aes(x=interval, y=avg_daily_steps))+
              geom_line()+
              
              theme(legend.position = "none",                                          axis.title.x = element_text("Interval"),axis.title.y=element_blank())+
              ggtitle("Average Daily Steps by Interval")+
              labs(caption = "Raw Data Plot, nulls removed")
              
```
Which interval has the highest average number of daily steps?

Using arrange on the data frame. Looks like 835 is the interval with the maximum daily steps.

```{r}
# arrange(raw_data_interval,desc(avg_daily_steps))
# head()
```
## Imputing missing values
 Calculate the total number of missing values coded as NA
```{r}
 summary(activity_raw_df)
```
Looks to be 2304 NA's in the steps column.

*Devise a strategy for filling in the missing values*

Going to use the activity_raw_df created earlier. This had the dates converted to a date field. Decided to use median steps to replace NA's.

```{r}
activity_cleaned_df <-activity_raw_df %>%
    mutate(steps=replace_na(steps,median(steps,na.rm = TRUE)))
summary(activity_cleaned_df)

```
Now put into same format as the steps df earlier.

``` {r}
clean_data_steps <- activity_cleaned_df %>% 
      select (date, steps, interval) %>%
      group_by(date,interval) %>% 
      summarise(total_daily_steps = sum(steps))

      summary(clean_data_steps)
```      

A pretty big change in median and mean steps once we replaced the NA's. Before, it was 10,765 for mean and median steps. As you can see above from the clean_data_steps summary, the mean and median have dropped to 9,354 and 10,395, respectively.

Now, for the histogram...
```{r}

  ggplot(data = clean_data_steps,mapping = aes(x=date, y=total_daily_steps, fill=total_daily_steps))+
              geom_bar(stat="identity")+
              scale_fill_continuous(low="light green", high = "green")+
              theme(legend.position = "none",                                          axis.title.x=element_blank(),axis.title.y=element_blank())+
              ggtitle("Total Steps by Day")+
              labs(caption = "Cleaned Data Plot")
```

## Are there differences in activity patterns between weekdays and weekends?

First things first, need to create a factor variable for weekday vs. weekend. I am going to modify the clean_data_steps data frame we just created with the imputed values for steps...

``` {r}
clean_data_steps_dow <- clean_data_steps %>% 
      mutate(dow = wday(date,label = TRUE)) %>% 
      mutate(dow_group = case_when(dow == "Sat"  ~ 'Weekend',
                                   dow == "Sun" ~ 'Weekend',
                                   dow == "Mon" ~ 'Weekday',
                                   dow == "Tue" ~ 'Weekday',
                                   dow == "Wed" ~ 'Weekday',
                                   dow == "Thu" ~ 'Weekday',
                                   dow == "Fri" ~ 'Weekday'))
```      
Now the weekday column is part of the df.  Let's do the plots..

```{r}
ggplot(data = clean_data_steps_dow,mapping = aes(x=interval, y=total_daily_steps, color = dow_group))+
              geom_line()+
              facet_wrap(~ dow_group)+
              
              theme(legend.position = "none",                                          axis.title.x = element_text("Interval"),axis.title.y=element_blank())+
              ggtitle("Total Daily Steps by Interval")+
              labs(caption = "Cleaned data")
              

```
Made 1 change so I could do a subsequent commit.







































